\documentclass{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{apacite}
\usepackage{caption}
\usepackage[shortlabels]{enumitem}

\begin{document}

\author{
  Jonsson, Haukur\\
  \texttt{11137304}
  \and
  Rajamanickam, Santhosh\\
  \texttt{11650702}
  \and
  Rapp, Max\\
  \texttt{11404310}
}
\title{DL Project Report}
\maketitle

\begin{enumerate}[(a)]
\item Redundant statements found:

\begin{itemize}
\item $VegitarianPizza$, $VegitarianPizzaEquivalent1$ and $VegitarianPizzaEquivalent2$ \\
All define the same pizzas in different ways

\item $SpicyPizzaEquivalent$ and $SpicyPizza$ \\
All define the same pizzas in different ways

\item $Caprina$ has $SundriedTomatoTopping$ which is a $TomatoTopping$ and $TomatoTopping$ as well. \\
Fix: Remove $TomatoTopping$ requirement.

\item $Giardiniera$ has $SlicedTomatoTopping$ which is a $TomatoTopping$ and $TomatoTopping$ as well. \\
Fix: Remove $TomatoTopping$ requirement.

\item $Pizza$ has $Food$ and $hasBase$ as $PizzaBase$ which is a subset of $Food$ \\
Fix: Remove $Food$ from definition of $Pizza$

\item $CheesyPizza$, $InterestingPizza$, $SpicyPizza$ and $Meaty Pizza$ have $Pizza$ in their definition but as $hasTopping$ has domain of $Pizza$, hence $Pizza$ is redundant. \\ 
Fix: Remove $Pizza$ from the definitions.

\item  $Hot$, $Medium$ and $Mild$ have $SubClass Of(Anonymous Ancestor)$ as "$Hot$ $or$ $Medium$ $or$ $Mild$" and also they have $Spiciness$ as subclass of where both means the same. \\
Fix: Remove "$Hot$ $or$ $Medium$ $or$ $Mild$" from $SubClass Of(Anonymous Ancestor)$ 

\end{itemize}

\item Inconsistent Pizza $\Rightarrow$ Consistent Pizza\\

Following are the reasons for the inconsistency and fixes we implemented to make the ontology consistent: \\

\begin{itemize}
\item Inconsistency: $IceCream$ had the property $hasTopping$ which had domain $Pizza$, which is disjoin with $IceCream$. \\
Fix: Removed $hasTopping$
\item Inconsistency: $CheesyVegetableTopping$ has top disjoint parents. \\
Fix: Deleted as the class is inconsistent and cannot exist.
\end{itemize}

After the above fixes the pizza ontology becomes consistent.

\item Extending the ontology with redundant statements

Below are some obvious redundant statement added to the definition of classes and verified by generating the inferred subsumptions
\begin{itemize}
\item Add $Food$ to the definition of $CheeseyPizza$, $InterestingPizza$ and $MeatyPizza$ which doesn’t change anything because it is the superclass/parent of $Pizza$ which is already in the definition.

\item Add $not$ $VegetarianPizza$ to the definition of $MeatyPizza$. This is redundant because $VegetarianPizza$ has $not$ $(hasTopping$ $some$ $MeatTopping)$ in its definition and $MeatyPizza$ has $(hasTopping$ $some$ $MeatTopping)$ in its definition.

\item Add $hasTopping$ $some$ $PepperTopping$ to subclass of $AmericanHot$. It is redundant as $PepperTopping$ is already implied by $AmericanHot$ being a subclass of "$hasTopping$ $some$ $HotGreenPepperTopping$ $and$ $hasTopping$ $some$ $JalapenoPepperTopping$" which are all subclasses/children of $PepperTopping$

\item Add $VegetarianTopping$ to the definition of $MushroomTopping$ and $OliveTopping$. This is redundant because $MushroomTopping$ and $OliveTopping$ are subclass of $VegetableTopping$ which is in turn $VegetarianTopping$

\end{itemize}











\end{enumerate}



\end{document}